Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-06-12 large_number_squeezing_with_atoms.tex
Normalise terminology.
I was using "fractional occupation of the non-zero momentum modes" in my bog section when the shorter phrase "non-condensed fraction" was used elsewhere.  I updated my section to use the shorter phrase.

We can understand the origin and scaling of this degradation of squeezing by considering the response of the condensate to small fluctuations about the mean field.  We do this using Bogoliubov theory \cite{PethickSmith}, which considers the first order quantum-mechanical fluctuations about the mean field.  To simplify the analysis we consider the case in which $U_{aa} \neq 0$, $U_{bb}=U_{ab} = 0$ and the condensate has a spatially-constant density in a box of side lengths $L_x$, $L_y$ and $L_z$.  In this limit the occupation of the non-zero momentum modes in the $\ket{a}$ internal state at the end of the pulse sequence is (a derivation is given in \ref{appendixBogDerivation})
\begin{align}
  n_a(\mathbf{k}, t_3) &= \expect{\hat{a}^\dagger(\mathbf{k},t) \hat{a}(\mathbf{k},t)} = \left[n_a \chi_{aa} \tau_\text{hold} \cos(\theta) \sinc\left(\omega_\mathbf{k} \tau_\text{hold}\right) \right]^2, \label{eqBogOccupation}
\end{align}
where the Bogoliubov mode frequency is $\omega_\mathbf{k} = \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \chi_{aa} n_a)}$, and the free particle frequency is $\omega^0_\mathbf{k} = \hbar \mathbf{k}^2/2 M$.  The non-zero momentum modes can be expected to have an impact on the squeezing of the system when their occupation is a non-negligible fraction of the total number of atoms.  The fractional occupation of these modes is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{1}{N}\sum_{\mathbf{k}\neq 0} n_a(\mathbf{k}, t), \label{eqCondensateDepletion}
\end{align}
where the sum over $\mathbf{k} \neq 0$ is taken over the available non-zero momentum modes $\mathbf{k} = (2\pi/L_x) n_x \hat{\mathbf{x}} + (2\pi/L_y) n_y \hat{\mathbf{y}} + (2\pi/L_z) n_z \hat{\mathbf{z}}$ with $n_x$, $n_y$ and $n_z$ arbitrary integers that are not all zero.

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/Bog_validation}
\caption{Plot of the fractional occupation of the non-zero momentum modes as a function of the hold time $\tau_\text{hold}$.  The blue dashed line is the Bogoliubov integral approximation given in Eq.~\ref{eqBogFraction1D}, the purple dot-dashed line is the Bogoliubov summation expression given in Eq.~\ref{eqCondensateDepletion}, and the solid black line is the result of a truncated Wigner simulation.  Parameters: $L=600\,\mu\text{m}$, $N=4 \times 10^5$, $n_a=n_b=2\times 10^5$, $\chi_{aa} = 2.67 \times 10^{-2}\, \text{s}^{-1}$, $\chi_{ab}=\chi_{bb}=0$, $\theta=0$.}
  \label{figBogSqueezingValidation}
\end{figure}

Figure~\ref{figBogSqueezingValidation} demonstrates the agreement between truncated Wigner simulations and the Bogoliubov theory for the fractional occupation of the non-zero momentum modes for parameters where the occupation of these modes approach 15\% of the total occupation of the system.  The disagreement for larger fractional occupations occurs because the Bogoliubov theory used in deriving \eqref{eqCondensateDepletion} is not number-conserving as it neglects the effect of the non-zero momentum modes on the condensate.

@@

To minimise the damaging effects of the non-zero momentum modes it is better to operate in a regime where the total occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ is small, which is also fortuitously the regime in which the Bogoliubov theory is a good approximation.  To better understand when $N_{\mathbf{k}\neq 0}/N$ will be small, we approximate the sum in \eqref{eqCondensateDepletion} with an integral.  In 1D Eq.~\eqref{eqCondensateDepletion} is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
We wish to approximate this sum by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k}, \label{eqCondensateDepletionFirstIntegralApproximation}
\end{align}
where $\Delta k = 2\pi/L$ and $L$ is the length of the 1D box condensate.  This approximation will be valid in the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ as for small $k$ the allowed values $k = n_x 2\pi /L$ will closely sample the region $\xi = \sqrt{\omega_k^0(\omega_k^0 + 2 \chi_{aa} n_a)} \tau_\text{hold} < \pi$ where $\sinc^2(\xi)$ is significant.

Changing integration variables to $\Phi = \omega^{0}_{\mathbf{k}}\tau_\text{hold}$ and defining $\Lambda_a = n_a \chi_{aa} \tau_\text{hold}$ the expression \eqref{eqCondensateDepletionFirstIntegralApproximation} can be written as
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 L \cos^2\theta } {\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} f^\text{1D}(\Lambda_a), \label{eqBogFraction1D}
\end{align}
where
\begin{align}
  f^\text{1D}(\Lambda) &= \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi.
\end{align}
In 2D and 3D the equivalent expressions are
\begin{align}
  \frac{N^{\text{2D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 A \cos^2\theta } {4\frac{\hbar}{M}\tau_\text{hold}} f^\text{2D}(\Lambda_a),  \label{eqBogFraction2D}\\
  \frac{N^{\text{3D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 V \cos^2\theta } {\sqrt{2}\pi^2\left(\frac{\hbar}{M}\tau_\text{hold}\right)^{3/2}} f^\text{3D}(\Lambda_a), \label{eqBogFraction3D}
\end{align}
where
\begin{align}
  f^\text{2D}(\Lambda) &= \int_0^{\infty} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi,\\
  f^\text{3D}(\Lambda) &= \int_0^{\infty} \sqrt{\Phi} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi,
\end{align}
and $A$ is the area of the two-dimensional system and $V$ is the volume of the three-dimensional system.  Figure~\ref{figBogSqueezingValidation} demonstrates the agreement in 1D between the integral approximation to $N_{\mathbf{k}\neq 0}/N$ in \eqref{eqBogFraction1D}, the summation expression in \eqref{eqCondensateDepletion} and a truncated Wigner simulation.

We wish to examine the scaling of the fractional occupation of the non-zero momentum modes $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate.  Other system parameters such as the total number $N$ and the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular this fixes the values of $\theta$, $\phi$ and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but $\tau_\text{hold}$ will vary to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The fractional occupation of the non-zero momentum modes can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The fractional occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the fractional occupation of these modes increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the fractional occupation of the non-zero momentum modes \emph{decreases} with volume.  This suggests that the best system to use will be a box potential, and the system size can be increased until the condensate depletion becomes negligibly small.

@@

Figure \ref{fig1D_squeezing_and_bog_scaling} and Figure \ref{fig2D_squeezing_and_bog_scaling} show the non-condensed fraction predicted by the Bogoliubov analysis compared to that predicted by a full numerical simulation, in one, two and three dimensions, as a function of the trap size. Also shown in the 1D and 2D cases is the maximum amount of number squeezing obtainable, as predicted by the numerical simulation. The full squeezing scaling curve is not shown in 3D, as for the smaller trap sizes the TWA method breaks down, and results are unreliable; for the larger trap sizes with negligible non-condensate fractions, however, the simulation was carried out and a relative number squeezing of 12.6dB was found. The traps are assumed to have a constant potential, which minimises the effects of the multimode issues described in section \ref{sec:MMdescription}.  We see that the squeezing does indeed degrade at large trap sizes in 1D, but remains constant in $2D$. For small trap sizes, the squeezing in 1D aymptotically approaches that of the ideal 0D system.

>>>>

We can understand the origin and scaling of this degradation of squeezing by considering the response of the condensate to small fluctuations about the mean field.  We do this using Bogoliubov theory \cite{PethickSmith}, which considers the first order quantum-mechanical fluctuations about the mean field.  To simplify the analysis we consider the case in which $U_{aa} \neq 0$, $U_{bb}=U_{ab} = 0$ and the condensate has a spatially-constant density in a box of side lengths $L_x$, $L_y$ and $L_z$.  In this limit the occupation of the non-zero momentum modes (the non-condensed fraction) in the $\ket{a}$ internal state at the end of the pulse sequence is (a derivation is given in \ref{appendixBogDerivation})
\begin{align}
  n_a(\mathbf{k}, t_3) &= \expect{\hat{a}^\dagger(\mathbf{k},t) \hat{a}(\mathbf{k},t)} = \left[n_a \chi_{aa} \tau_\text{hold} \cos(\theta) \sinc\left(\omega_\mathbf{k} \tau_\text{hold}\right) \right]^2, \label{eqBogOccupation}
\end{align}
where the Bogoliubov mode frequency is $\omega_\mathbf{k} = \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \chi_{aa} n_a)}$, and the free particle frequency is $\omega^0_\mathbf{k} = \hbar \mathbf{k}^2/2 M$.  The non-zero momentum modes can be expected to have an impact on the squeezing of the system when their occupation is a non-negligible fraction of the total number of atoms.  The fractional occupation of these modes is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{1}{N}\sum_{\mathbf{k}\neq 0} n_a(\mathbf{k}, t), \label{eqCondensateDepletion}
\end{align}
where the sum over $\mathbf{k} \neq 0$ is taken over the available non-zero momentum modes $\mathbf{k} = (2\pi/L_x) n_x \hat{\mathbf{x}} + (2\pi/L_y) n_y \hat{\mathbf{y}} + (2\pi/L_z) n_z \hat{\mathbf{z}}$ with $n_x$, $n_y$ and $n_z$ arbitrary integers that are not all zero.

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/Bog_validation}
\caption{Plot of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ as a function of the hold time $\tau_\text{hold}$.  The blue dashed line is the Bogoliubov integral approximation given in Eq.~\ref{eqBogFraction1D}, the purple dot-dashed line is the Bogoliubov summation expression given in Eq.~\ref{eqCondensateDepletion}, and the solid black line is the result of a truncated Wigner simulation.  Parameters: $L=600\,\mu\text{m}$, $N=4 \times 10^5$, $n_a=n_b=2\times 10^5$, $\chi_{aa} = 2.67 \times 10^{-2}\, \text{s}^{-1}$, $\chi_{ab}=\chi_{bb}=0$, $\theta=0$.}
  \label{figBogSqueezingValidation}
\end{figure}

Figure~\ref{figBogSqueezingValidation} demonstrates the agreement between truncated Wigner simulations and the Bogoliubov theory for the non-condensed fraction for parameters where the occupation of these modes approach 15\% of the total occupation of the system.  The disagreement for larger non-condensed fractions occurs because the Bogoliubov theory used in deriving \eqref{eqCondensateDepletion} is not number-conserving as it neglects the effect of the non-zero momentum modes on the condensate.

@@

To minimise the damaging effects of the non-condensed fraction it is better to operate in a regime where the total occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ is small, which is also fortuitously the regime in which the Bogoliubov theory is a good approximation.  To better understand when $N_{\mathbf{k}\neq 0}/N$ will be small, we approximate the sum in \eqref{eqCondensateDepletion} with an integral.  In 1D Eq.~\eqref{eqCondensateDepletion} is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
We wish to approximate this sum by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k}, \label{eqCondensateDepletionFirstIntegralApproximation}
\end{align}
where $\Delta k = 2\pi/L$ and $L$ is the length of the 1D box condensate.  This approximation will be valid in the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ as for small $k$ the allowed values $k = n_x 2\pi /L$ will closely sample the region $\xi = \sqrt{\omega_k^0(\omega_k^0 + 2 \chi_{aa} n_a)} \tau_\text{hold} < \pi$ where $\sinc^2(\xi)$ is significant.

Changing integration variables to $\Phi = \omega^{0}_{\mathbf{k}}\tau_\text{hold}$ and defining $\Lambda_a = n_a \chi_{aa} \tau_\text{hold}$ the non-condensed fraction \eqref{eqCondensateDepletionFirstIntegralApproximation} can be written as
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 L \cos^2\theta } {\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} f^\text{1D}(\Lambda_a), \label{eqBogFraction1D}
\end{align}
where
\begin{align}
  f^\text{1D}(\Lambda) &= \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi.
\end{align}
In 2D and 3D the equivalent expressions are
\begin{align}
  \frac{N^{\text{2D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 A \cos^2\theta } {4\frac{\hbar}{M}\tau_\text{hold}} f^\text{2D}(\Lambda_a),  \label{eqBogFraction2D}\\
  \frac{N^{\text{3D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 V \cos^2\theta } {\sqrt{2}\pi^2\left(\frac{\hbar}{M}\tau_\text{hold}\right)^{3/2}} f^\text{3D}(\Lambda_a), \label{eqBogFraction3D}
\end{align}
where
\begin{align}
  f^\text{2D}(\Lambda) &= \int_0^{\infty} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi,\\
  f^\text{3D}(\Lambda) &= \int_0^{\infty} \sqrt{\Phi} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi,
\end{align}
and $A$ is the area of the two-dimensional system and $V$ is the volume of the three-dimensional system.  Figure~\ref{figBogSqueezingValidation} demonstrates the agreement in 1D between the integral approximation to $N_{\mathbf{k}\neq 0}/N$ in \eqref{eqBogFraction1D}, the summation expression in \eqref{eqCondensateDepletion} and a truncated Wigner simulation.

We wish to examine the scaling of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate.  Other system parameters such as the total number $N$ and the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular this fixes the values of $\theta$, $\phi$ and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but $\tau_\text{hold}$ will vary to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The non-condensed fraction can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The non-condensed fraction $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the non-condensed fraction increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the non-condensed fraction \emph{decreases} with volume.  This suggests that the best system to use will be a three-dimensional box potential, and the system size can be increased until the condensate depletion becomes negligibly small.

@@

Figures~\ref{fig1D_squeezing_and_bog_scaling}--\ref{fig3D_Bog_mode_occupation_scaling} show the non-condensed fraction predicted by the Bogoliubov analysis compared to that predicted by a full numerical simulation, in one, two and three dimensions, as a function of the trap size. Also shown in the 1D and 2D cases is the maximum amount of number squeezing obtainable, as predicted by the numerical simulation. The full squeezing scaling curve is not shown in 3D, as for the smaller trap sizes the TWA method breaks down, and results are unreliable; for the larger trap sizes with negligible non-condensate fractions, however, the simulation was carried out and a relative number squeezing of 12.6dB was found. The traps are assumed to have a constant potential, which minimises the effects of the multimode issues described in section \ref{sec:MMdescription}.  We see that the squeezing does indeed degrade at large trap sizes in 1D, but remains constant in $2D$. For small trap sizes, the squeezing in 1D aymptotically approaches that of the ideal 0D system.
